A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2015-09-29 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Commit Annika's changes

%Begin with a focus on more recent stuff such as Gödel, not Anselm and so on. Focus on the improved level of formalization of the argument through Leibniz/Descartes/Kant and so on. 
%Think about how much can be claimed about Russell, Descartes, Kant and so on... 
%Russell eliminated or expained as a valuable contribution to the formal proof. 
%check the Solmon paper to se about proofs in KB. Other ontological proofs. 
%Relate introduction to the beginning quote
%spell check
%How to use the words God, God-like, god-like and greatest conceivable being. 
%Theorem reference missing in the proof of modal collapse
%Write the section on axiom B and the strength of the modal system that is necessary.

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considers a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence.

A major critique of this argument is that we do not know whether the concept of greatest conceivable being in fact designates anything or if it is inconsistent, like a round square \citep{fitting}[p.134].  
%The question is if the argument is sound. 
As Bertrand Russell pointed out, the definition of greatest conceivable being allows us to define properties, like having boots, which the being then also must have \citep{citation needed}.

Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. The strongest claim that can be proven by the ontological argument is a conditional claim of necessary existence: if God exists, then God necessarily exists \citep{citation needed}.

St. Anselm's argument was further elaborated by Descartes and Leibniz. Leibniz identified the possible existence of God as a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This means that it is possible to satisfy all perfections at once which implies that 
the existence of a greatest conceivable being with all these properties is possible.

G\"odel studied Leibniz's work \citep{Adams} and brought his ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being having all positive properties.

G\"odel's notion of positive property and Leibniz's notion of perfection differ \citep{Adams}. The main formal distinction seems to be that G\"odel's positive properties are not just atomic properties, like Leibniz's perfections, but can also consist of complex combinations of atomic properties \citep{fitting}[p.139.] (TODO: Check this claim). In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being God-like is positive. While this step is intuitively and informally clear, it is not easily formalizable in a standard logical calculus, because it requires inferring that being God-like is a (possibly infinite) conjunction of positive properties, from the universal definition of God-likeness as the property satisfied by individuals who have all positive properties. This interplay between universal quantification (in the definition of a God-like being) and in!
 finite conjunctions (in G\"odel's axiom for positive properties) is a technical difficulty that probably explains why, starting with Scott \cite{Scott}, to whom G\"odel confided his manuscript, this axiom of G\"odel's has been replaced by another that simply assumes the positivity of the property of being God-like.

The main criticism against G\"odel's ontological argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. Many recent works \cite{todo} on the ontological argument have focused on proposing modifications of the argument that would not entail a modal collapse. This is discussed in greater detail in Section \ref{sec:collapse}, but possible solutions are not implemented in this paper.

%This axiom is unfortunately equivalent to the possibility of God's existence given the other axioms, leading to criticisms that it might be too strong \citep{sobel2}\citep{fitting}[Section 11.4. Objections.].

%The possibility  God's existence in turn is equivalent to  God's necessary existence as well as  God's existence itself. Thus, we have an axiom which seems to be equivalent to the conclusion.

 % This axiom of the positiveness of conjunctions can be given a third order formulation which is the reason for the claim that third-order quantification may be required \citep{and}. %\citep{fitting}[p. 148.]

% Another, possibly undesirable, consequence of the formalism is that there can only be one God, which is defined by the axioms. Namely, if one assumes that the theory contains the equality relation then a God has the property of being identical to itself. Since any other god-like being would have at least the same properties it shares the property of being identical to the one God. Thus, we have proved monotheism \citep{sobel2}[Ch. 4, section 3.3.2]. %\citep{fitting}[The proof is Exercise 7.1, p.163].

The aim of this paper is to present two detailed formal proofs of G\"odel's ontological argument \cite{Goedel} in a natural deduction calculus (as defined in Section \ref{sec:calculus}). The natural deduction style \citep{gentzen,prawitz} was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus. The second contribution of the paper is a new proof (also in natural deduction style) of the lemmas, theorems and corollaries in G\"odel's manuscript. In contrast to Scott's proofs \cite{Scott}, the proofs presented here are simpler and shorter, as discussed in Section \ref{sec:conclusion}.

@@

ToDo: decide which connectives and corresponding inference rules will be primitive.

@@

ToDo: decide if rules for diamond should be added.

@@

\subsection{Suitability for Rigid Higher-Order Modal Logic K}

@@

\section{Some Derivable Modal Principles}
\label{sec:AxiomB}.

As mentioned in \citep{and90}[Footnote 5], there are sources that discuss what modal principles are needed for the ontological argument. The standard system for the argument is S5. Through an analysis of the proofs weaker systems, in particular KB, can be shown to be sufficient for versions of the ontological argument \citep{sobel2}[p. 152]. The proof presented in Section \ref{sec:newproof} is such a version and explicitly shows a single use of Brouwer Reduction Principle, which, in turn, is provable from axiom B. All other proof steps are formalizable in a basic second-order modal logic K.

In standard first-order modal logics the modal axioms correspond to geometric conditions on the frames of Kripke models \citep{negri}. The modal axiom {\bf T} corresponds to reflexivity of the frame and {\bf B} corresponds to symmetry of the frame. The frames of the modal logic {\bf S5} are in addition transitive.

%G\"odel mentions in the notes which are saved of his proof that the modal system required for proving his theorem is a second order version of the modal system {\bf S5}. In \citep{and90} it is conjectured that the principles of proof could be restricted to the weaker modal logic {\bf KB}. The modal logic {\bf KB} consists of the basic modal system {\bf K} and the modal axioms {\bf T} and {\bf B}.

@@

To prove G\"odel's ontological proof in natural deduction we use a modal distribution principle derivable in K and  Brouwer's reduction theorem derivable from the modal Axiom B.

@@

\subsection{Axiom B and Brouwer's Reduction Principle}

Brouwer's reduction theorem is a version of axiom B for modal logic and a property that enforces symmetry as a geometric property for the frame of a Kripke sematics.

@@

ToDo: Is this principle really commonly called ``S5 Iteration Principle''?

@@

\begin{prooftree}
\AXC{ToDo}
\end{prooftree}

@@

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if the exemplification of the god-likeness property is possible, then it is necessary and a proof that the exemplification is in fact possible.

@@

\begin{theorem}
\label{T2}
Being God-like is an essence of any God-like being:
$$
\all x. G(x) \imp \ess{G}{x}
$$
\end{theorem}
\begin{proof}

@@

\begin{lemma}
\label{L2}
If the existence of a God-like being is possible, then it is necessary:
$$
\pos \ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

A outline of the proof is that firstly all positive properties are proven to be possibly exemplified (Theorem \ref{T1}). Then this theorem is applied to the property of god-likeness, thus showing the possible existence of a god-like being (Corollary \ref{C1}). In Lemma \ref{L1} it is classically proven that the hypothetically assumed existence of a god-like being implies that the existence is necessary. Lastly, the necessary existence of a god-like being is proven (Theorem \ref{T2}) from Corollary \ref{C1} and Lemma \ref{L1} by properties of the modal system. The properties required for the proof are the distribution principle and Brouwer's reduction theorem mentioned in Section \ref{sec:AxiomB}. The proof of Theorem \ref{T2} follows the proof of Sobel \citep{sobel2}[p. 126--127].

@@

The exemplification of the god-likeness property is provable from its possible exemplification by the modal distribution principle \ref{DP} and Brouwer's reduction principle \ref{BRP}. The necessary exemplification of the god-likeness property can then be deduced from lemma \ref{L1}. This single use of the modal distribution principle \ref{DP} and Brouwer's reduction principle \ref{BRP} follows the presentation of in Sobel \citep{sobel2}[p.126--127].

@@

%Is the distribution principle and brouwers Red thm better than using axiom B straight? Check previous versions... ***

%From that it is possible to that it exists with modal inferences... **** and axioms...

@@

Anderson's solution \citep{citation needed} modifies the definitions of god-like being and essence, and eliminates half of an axiom. This not only avoids the modal collapse, but also makes two of G\"odel's five axioms derivable from the others \citep{hajek} under some implicit additional assumptions \citep{fuhrmann}. Another solution involving more substantial modifications is that of Bj{\o}rdal \citep{bjordal,fuhrmann}.

@@

@@

@@

\bibitem[{\itshape Goedel(***)}]{Goedel} 
G\"odel, K. {\itshape Appendix A. Notes in Kurt G\"odel's hand} in Logic and Theism: Arguments for and against Beliefs in God by Sobel, J. H.

@@

@@

\bibitem[{\itshape Sobel(1987)}]{sobel}
Sobel, J. H. 1987. {\itshape G\"odel's Ontological Proof}. In: edited by J. J. Thompson. {\itshape On being and saying : essays for Richard Cartwright},  MIT Press.

@@

***Gentzen and Prawitz: references to Natural Deduction. 
Adams, Fuhrmann, Hajek and Bjordal: References to amendations. 
Church: Reference to simple type theory.

>>>>

%spell check

%Extra:

%Difference in the formalizations of Leibniz and Gödel? Why did Gödel change to conjunction?

%Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. The strongest claim that can be proven by the ontological argument is a conditional claim of necessary existence: if God exists, then God necessarily exists \citep{citation needed}.

% Another, possibly undesirable, consequence of the formalism is that there can only be one God, which is defined by the axioms. Namely, if one assumes that the theory contains the equality relation then a God has the property of being identical to itself. Since any other god-like being would have at least the same properties it shares the property of being identical to the one God. Thus, we have proved monotheism \citep{sobel2}[Ch. 4, section 3.3.2]. %\citep{fitting}[The proof is Exercise 7.1, p.163].  Inclusion of identity relation gives monothemism?*** Reference missing?

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considers a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence. St. Anselm's argument was further elaborated by Descartes, Leibniz and Kant.

Leibniz identified the possible existence of God as a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This means that it is possible to satisfy all perfections at once which implies that the existence of a greatest conceivable being with all these properties is possible.

G\"odel bulit on Leibniz's work \cite{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \cite{Goedel} as well as notes by Scott \cite{Scott}, in whom he confided his proof.

The increase in formality of the ontological argument has required a development of the basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being god-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a god-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{Scott}, this axiom o!
 f G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

The main criticism against the ontological argument is that its soundness has been doubted. As Fitting \citep{fitting}[p.134] described the problem ''we do not know whether the concept of greatest conceivable being in fact designates anything or if it is inconsistent, like a round square''.

The main criticism against G\"odel's formal argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. Many recent works on the ontological argument have focused on proposing modifications of the argument that would not entail a modal collapse. This is discussed in greater detail in Section \ref{sec:collapse}, but possible solutions are not implemented in this paper.

The aim of this paper is to present two detailed formal proofs of G\"odel's ontological argument in a natural deduction calculus (as defined in Section \ref{sec:calculus}). For a comprehensive introduction to natural deduction the reader can consult \citep{prawitz}. The natural deduction style was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus. In section \ref{sec:compl-sound} we prove that the calculus used is sound and complete relative to an axiomatic modal calculus and this detailed formalization of Scott's version has as been formally verified. The second contribution of the paper is a new proof (also in natural deduction style) in a natural deduction calculus. In contrast to Scott's proof \cite{Scott} formalized in the modal calculus S5, the latter proof is formalized in KB, as discussed in Section \ref{sec:AxiomB}.

@@

@@

@@

\subsection{Suitability for Rigid Higher-Order Modal Logic K}\label{sec:compl-sound}

@@

\section{Derivable Modal Principles}
\label{sec:AxiomB}.

In standard first-order modal logics the modal axioms correspond to geometric frame properties for Kripke models \cite{negri}. The modal axiom {\bf T} corresponds to reflexivity of the frame and {\bf B} corresponds to symmetry of the frame. The frames of the modal logic {\bf S5} are in addition transitive.

There are discussion of what modal principles are needed for the ontological argument. 
G\"odel mentions in his notes that the modal system required for proving his theorem is a second order version of the modal system {\bf S5}. In \cite{and90} it is conjectured that the principles of proof could be restricted to the weaker modal logic {\bf KB}. The modal logic {\bf KB} consists of the basic modal system {\bf K} and the modal axioms {\bf T} and {\bf B}.

Through an analysis of the proofs weaker systems, in particular KB, can be shown to be sufficient for versions of the ontological argument \citep{sobel2}[p. 152]. The proof presented in Section \ref{sec:newproof} is such a version and explicitly shows a single use of Brouwer Reduction Principle, which, in turn, is provable from axiom B. All other proof steps are formalizable in a basic second-order modal logic K.

We will prove the necessary modal principles as two lemmas; a modal distribution principle derivable in K and Brouwer's reduction principle derivable in KB. In addition we prove an iteration principle, which is valid in S5, for the proof in section \ref{sec:scott}. The iteration principle is derived in KB with axiom 5.

@@

@@

\subsection{Brouwer's Reduction Principle}

@@

@@

Axiom 5 is $\pos A\imp \nec \pos A$.

\AXC{Brouwer's reduction principle}\dashedLine
\UIC{$\pos \nec A\imp A$}\RightLabel{$\nec_I$}
\UIC{$\nec(\pos \nec A\imp A)$}
\AXC{Axiom K}\dashedLine
\UIC{$\nec(\pos \nec A\imp A)\imp (\nec\pos \nec A\imp \nec A)$}\RightLabel{$\imp_E$}
\BIC{$\nec\pos \nec A\imp \nec A$}

\AXC{Axiom 5 on $\nec A$}\dashedLine
\UIC{$\pos \nec A\imp \nec \pos \nec A$}

\AXC{$ $}\RightLabel{$1$}
\UIC{$\pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec \pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec A$}\RightLabel{$\imp_I^1$}
\UIC{$\pos \nec A \imp  \nec A$}
\end{prooftree}
\end{small}

@@

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The following proof is a detailed formalization of Scott's proof that has been verified in Coq \cite{***}. The formal argument has  two parts; a proof that if the exemplification of the god-likeness property is possible, then it is necessary and a proof that the exemplification is in fact possible.

@@

\begin{theorem}
\label{T2}
Being god-like is an essence of any god-like being:
$$
\all x. G(x) \imp \ess{G}{x}
$$
\end{theorem}
\begin{proof}

@@

\begin{lemma}
\label{L2}
If the existence of a god-like being is possible, then it is necessary:
$$
\pos \ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

@@

\begin{theorem}
\label{T3}
Necessarily, there exists a god-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

@@

In this section we present a new proof in KB. The outline of the proof is: 
Firstly all positive properties are proven to be possibly exemplified (Theorem \ref{T1}). Then this theorem is applied to the property of god-likeness, thus showing the possible existence of a god-like being (Corollary \ref{C1}). In Lemma \ref{L1} it is classically proven that the hypothetically assumed existence of a god-like being implies that the existence is necessary. Lastly, the necessary existence of a god-like being is proven (Theorem \ref{T2}) from Corollary \ref{C1} and Lemma \ref{L1} by properties of the modal system. The properties required for the proof are the distribution principle and Brouwer's reduction theorem mentioned in Section \ref{sec:AxiomB}. The proof of Theorem \ref{T2} follows the proof of Sobel \citep{sobel2}[p. 126--127].

@@

The exemplification of the god-likeness property is provable from its possible exemplification by the modal distribution principle \ref{DP} and Brouwer's reduction principle \ref{BRP}. The necessary exemplification of the god-likeness property can then be deduced from lemma \ref{L1}. This single use of the modal distribution principle \ref{DP} and Brouwer's reduction principle \ref{BRP} follows the presentation of Sobel \citep{sobel2}[p.126--127].

@@

@@

Anderson's solution \citep{and***} modifies the definitions of god-like being and essence, and eliminates half of an axiom. This not only avoids the modal collapse, but also makes two of G\"odel's five axioms derivable from the others \citep{hajek} under some implicit additional assumptions \citep{fuhrmann}. Another solution involving more substantial modifications is that of Bj{\o}rdal \cite{bjordal,fuhrmann}.

@@

\bibitem[{\itshape Adams(1995)}]{adams}
Adams, R. M. 1995. Introductory note to [G\"odel 1970].

@@

\bibitem[{\itshape Bj{\o}rdal(1999)}]{bjordal}
Bj{\o}rdal, F. 1999. 
Understanding G\"odel's Ontological Argument, in Timothy Childers (ed.) The Logica Yearbook 1998, pp. 214-217, Filosofia.

\bibitem[{\itshape Church(1940)}]{church}
Church, A. 1940. {\itshape A Formulation of the Simple Theory of Types}, Journal of Symbolic Logic, 5: 56?68.

@@

\bibitem[{\itshape Fuhrmann(2005)}]{fuhrmann}
Fuhrmann, A. 2005.
Existenz und Notwendigkeit -- Kurt G\"odel's Axiomatische Theologie, in Olsson e.J., Schr\"oder-Heister, P. and Spohn, W. (eds.) Logik in der Philosophie, pp. 349-374, Synchr.-Wissenschafts-Verlag.

\bibitem[{\itshape Goedel(1970)}]{Goedel} 
G\"odel, K. Ontological Proof. In Kurt G\"odel Collected Works vol. III. Ed. Feferman et al., pp. 403--404, 139, 145 or {\itshape Appendix A. Notes in Kurt G\"odel's hand} in [Sobel 2001].

\bibitem[{\itshape Hajek(1996)}]{hajek}
Hajek, P. 1996. Magari and others on G\"odel's Ontological Proof, in Ursini et alii
(eds.) Logic and Logical Algebra, pp. 125-136, Marcel Dekker.

@@

\bibitem[{\itshape Prawitz (2006)}]{prawitz}
Prawitz, D. 2006 (1st ed. 1965). {\itshape Natural deduction: a proof-theoretical study}. Mineola, New York: Dover publications.

@@

\bibitem[{\itshape Sobel(1987)}]{sobel}
Sobel, J. H. 1987. {\itshape G\"odel's Ontological Proof}. In: J. J. Thompson (ed.). {\itshape On being and saying : essays for Richard Cartwright},  MIT Press.

@@

***Adams, Fuhrmann, Hajek and Bjordal: Confirm references to ammendations and Church. 
****References for \cite{coq} and \cite{isabelle} and other papers checked step-by-step in Coq \cite{BenzmullerWolzenlogel-Paleo}. Reference to the inspiration for the modal rules. 
***One Anderson reference incomplete***
***Scott reference: the year is missing***
